---
title: 'P&S-2022: Lab assignment 2'
author: "Name1, Name2, Name3"
output:
  html_document:
    df_print: paged
---

## 	General comments and instructions
*  Complete solution will give you $\bf 4$ points (out of 100 total). Submission deadline is **23:59 of 06 November 2022**  
*  The report must be prepared as an _R notebook_; you must submit to **cms** both the source _R notebook_ **and** the generated html file  
*  At the beginning of the notebook, provide a work-breakdown structure estimating efforts of each team member  
*  For each task, include 
    +  problem formulation and discussion (what is a reasonable answer to discuss);  
    +  the  corresponding $\mathbf{R}$ code with comments (usually it is just a couple of lines long);  
    +  the statistics obtained (like sample mean or anything else you use to complete the task) as well as histograms etc to illustrate your findings;  
    +  justification of your solution (e.g. refer to the corresponding theorems from probability theory);  
    +  conclusions (e.g. how reliable your answer is, does it agree with common sense expectations etc)  
*  The __team id number__ referred to in tasks is the __two-digit__ ordinal number of your team on the list. Include the line __set.seed(team id number)__ at the beginning of your code to make your calculations reproducible. Also observe that the answers **do** depend on this number!  
*  Take into account that not complying with these instructions may result in point deduction regardless of whether or not your implementation is correct. 

### Task 1 (Nazar Demchuk)

#### In this task, we discuss the $[7,4]$ Hamming code and investigate its reliability. That coding system can correct single errors in the transmission of $4$-bit messages and proceeds as follows:

-   given a message $\mathbf{m} = (a_1 a_2 a_3 a_4)$, we first encode it to a $7$-bit *codeword* $\mathbf{c} = \mathbf{m}G = (x_1 x_2 x_3 x_4 x_5 x_6 x_7)$, where $G$ is a $4\times 7$ *generator* matrix\
-   the codeword $\mathbf{c}$ is transmitted, and $\mathbf{r}$ is the received message\
-   $\mathbf{r}$ is checked for errors by calculating the *syndrome vector* $\mathbf{z} := \mathbf{r} H$, for a $7 \times 3$ *parity-check* matrix $H$\
-   if a single error has occurred in $\mathbf{r}$, then the binary $\mathbf{z} = (z_1 z_2 z_3)$ identifies the wrong bit no. $z_1 + 2 z_2 + 4z_3$; thus $(0 0 0)$ shows there was no error (or more than one), while $(1 1 0 )$ means the third bit (or more than one) got corrupted\
-   if the error was identified, then we flip the corresponding bit in $\mathbf{r}$ to get the corrected $\mathbf{r}^* = (r_1 r_2 r_3 r_4 r_5 r_6 r_7)$;\
-   the decoded message is then $\mathbf{m}^*:= (r_3r_5r_6r_7)$.

#### The **generator** matrix $G$ and the **parity-check** matrix $H$ are given by

$$  
    G := 
    \begin{pmatrix}
        1 & 1 & 1 & 0 & 0 & 0 & 0 \\
        1 & 0 & 0 & 1 & 1 & 0 & 0 \\
        0 & 1 & 0 & 1 & 0 & 1 & 0 \\
        1 & 1 & 0 & 1 & 0 & 0 & 1 \\
    \end{pmatrix},
 \qquad 
    H^\top := \begin{pmatrix}
        1 & 0 & 1 & 0 & 1 & 0 & 1 \\
        0 & 1 & 1 & 0 & 0 & 1 & 1 \\
        0 & 0 & 0 & 1 & 1 & 1 & 1
    \end{pmatrix}
$$

#### Assume that each bit in the transmission $\mathbf{c} \mapsto \mathbf{r}$ gets corrupted independently of the others with probability $p = \mathtt{id}/100$, where $\mathtt{id}$ is your team number. Your task is the following one.

1.  Simulate the encoding-transmission-decoding process $N$ times and find the estimate $\hat p$ of the probability $p^*$ of correct transmission of a single message $\mathbf{m}$. Comment why, for large $N$, $\hat p$ is expected to be close to $p^*$.\
2.  By estimating the standard deviation of the corresponding indicator of success by the standard error of your sample and using the CLT, predict the \emph{confidence} interval $(p^*-\varepsilon, p^* + \varepsilon)$, in which the estimate $\hat p$ falls with probability at least $0.95$.\
3.  What choice of $N$ guarantees that $\varepsilon \le 0.03$?\
4.  Draw the histogram of the number $k = 0,1,2,3,4$ of errors while transmitting a $4$-digit binary message. Do you think it is one of the known distributions?

#### You can (but do not have to) use the chunks we prepared for you

#### First, we set the **id** of the team and define the probability $p$ and the generator and parity-check matrices $G$ and $H$

```{r}
# your team id number 
                          ###
id <- 28                  ### Change to the correct id!
                          ###
set.seed(id)
p <- id/100
# matrices G and H
G <- matrix(c(1, 1, 1, 0, 0, 0, 0,
		1, 0, 0, 1, 1, 0, 0,
		0, 1, 0, 1, 0, 1, 0,
		1, 1, 0, 1, 0, 0, 1), nrow = 4, byrow = TRUE)
H <- t(matrix(c(1, 0, 1, 0, 1, 0, 1,
		0, 1, 1, 0, 0, 1, 1,
		0, 0, 0, 1, 1, 1, 1), nrow = 3, byrow = TRUE))
cat("The matrix G is: \n")
G
cat("The matrix H is: \n")
H
cat("The product GH must be zero: \n")
(G%*%H) %%2
```

#### Next, generate the messages

```{r}
# generate N messages
message_generator <- function(N) {
  matrix(sample(c(0,1), 4*N, replace = TRUE), nrow = N)
}
# using a large number of entries to get the proper results
N = 10000
messages <- message_generator(N)
# multiplying randomly generated messages matrix and G matrix
codewords <- (messages %*% G) %% 2 # codewords mod 2
cat("The matrix of encoded messages is: \n")
head(codewords)
```

#### Generate random errors; do not forget that they occur with probability $p$! Next, generate the received messages

```{r}
# generating random errors
# changeBit <- function(x) {
# 	return (sample(c(x, 1 - x), size=1, prob=c(1 - p, p)))
# }
errors_generator <- function(N) {
	matrix(sample(c(0,1), 7*N, replace = TRUE, prob=c(1-p, p)), nrow=N)
}
errors <- errors_generator(N)
cat("The generated errors matrix is: \n")
head(errors)
modulo2 <- function(x) {
	return (x %% 2);
}
received <- codewords + errors
received <- t(apply(received, 1, modulo2))
cat("The matrix of received words with possible errors is: \n")
head(received)
```

The next steps include detecting the errors in the received messages, correcting them, and then decoding the obtained messages. After this, you can continue with calculating all the quantities of interest

```{r}
decode <- function(x) {
	# multiplying matrices
	z <- x %*% H
	z <- z %% 2
	# checking index of the corrupted bit (if exists)
	idx <- z[1,1] + 2 * z[1,2] + 4 * z[1,3]
	# correcting the message (flipping the bit)
	if (idx > 0) {
		x[idx] <- (x[idx] + 1) %% 2
	}
	# return the decoded message
	return(c(x[3],x[5],x[6],x[7]))
}
decoded_messages <- t(apply(received, 1, decode))
cat("The matrix of decoded messages is: \n")
head(decoded_messages)
```

```{r}
actual_prob <- sum(
	decoded_messages[,1]==messages[,1]& 
	decoded_messages[,2]==messages[,2]&
	decoded_messages[,3]==messages[,3]&
	decoded_messages[,4]==messages[,4]) / N
cat("The actual probability is: \n")
actual_prob
cat("The hypothetical probability is: \n")
hypothetical_prob <- (1-p)**7 + 7*p*(1-p)**6
hypothetical_prob
cat("The standard deviation is: \n")
x <- sqrt(0.3733617*(1-0.3733617)**2+(1-0.3733617)*(0-0.3733617)**2)
x
```

The probability $p*$ is equal to the probability that there were no digits corrupted (so the message is automatically correct) or there was a single corruption in all of the digits transmitted (so we can correct it using our approach). So, $p = (1-p)^7 + {7\choose 1}(1-p)^6p=(\frac{72}{100})^7 + 7\cdot(\frac{72}{100})^6*(\frac{28}{100})=0.3733617$.

Here we can say that because we used a fairly large amount of data (as we generated \$N = 10000\$ messages), the actual $\hat p$ that we got is expected to be close to $p^*$ (by the Law of Large Numbers).

First, we have estimate the standard deviation of the corresponding indicator of success by the standard error of our sample. So, the standard deviation can be calculated by finding each deviation from its expected value, squaring it, multiplying it by its probability, adding the products, and then taking the square root. So,

$$\sigma^2=0.3733617(1-0.3733617)^2+0.6266383(0-0.3733617)^2=0.2339627$$

which gives us $\sigma=0.483697$.

Let's assume that $\sigma\approx0.48$ and $p^*\approx0.37$.

By the Central Limit Theorem, we can say that $Z=\frac{\sqrt{N}}{0.48}(\hat{p}-0.37)$ is a random variable with a standard normal distribution.

Our interval $(p^*-\epsilon, p^*+\epsilon)$ has to make our estimate $\hat p$ fall with probability of at least $0.95$. So, $P(p^*-\epsilon < \hat{p}<p^*+\epsilon)=0.95$. From this, we get that $2\Phi(\frac{\epsilon\sqrt{N}}{0.38})-1=0.95$ and $\Phi(\frac{\epsilon\sqrt{N}}{0.48}) = 0.975$.

So, the value of our $\frac{\epsilon\sqrt{N}}{0.48}$ is $1.95996398454$.

$\frac{\epsilon\sqrt{N}}{0.48}\approx1.96, \epsilon=\frac{0.9408}{\sqrt{N}}$

If $N = 10000$ (as it was the value used in our experiment), then $\epsilon\approx0.009408$.

If the condition is that $\epsilon \le 0.03$, then $N \ge (\frac{0.9408}{0.03})^2 \approx 983.4496$. That means that $N \ge 984$.

```{r}
# checking if the message was decoded properly
correct_messages_truth_table <- decoded_messages!=messages
head(correct_messages_truth_table)
# getting the sum of incorrect messages for every bit
correct_messages_truth_table <- apply(correct_messages_truth_table, 1, sum)
hist(x=correct_messages_truth_table,
     main="The Amount of Corrupted Bits",
     xlab="Corruptions",
     ylab="Probability")
```

The fact that the column with $0$ errors in the message is the highest is not surprising because there was only a $28$ percent chance of every given bit being corrupted (which is still quite high, though). Also, for $k=1,2,3$ the graph is pretty similar to the binomial distribution graph.


### Task 2 (Dmytro Hrebeniuk)

#### In this task, we discuss a real-life process that is well modelled by a Poisson distribution. As you remember, a Poisson random variable describes occurrences of rare events, i.e., counts the number of successes in a large number of independent random experiments. One of the typical examples is the **radioactive decay** process.

#### Consider a sample of radioactive element of mass $m$, which has a big *half-life period* $T$; it is vitally important to know the probability that during a one second period, the number of nuclei decays will not exceed some critical level $k$. This probability can easily be estimated using the fact that, given the *activity* ${\lambda}$ of the element (i.e., the probability that exactly one nucleus decays in one second) and the number $N$ of atoms in the sample, the random number of decays within a second is well modelled by Poisson distribution with parameter $\mu:=N\lambda$. Next, for the sample of mass $m$, the number of atoms is $N = \frac{m}{M} N_A$, where $N_A = 6 \times 10^{23}$ is the Avogadro constant, and $M$ is the molar (atomic) mass of the element. The activity of the element, $\lambda$, is $\log(2)/T$, where $T$ is measured in seconds.

#### Assume that a medical laboratory receives $n$ samples of radioactive element ${{}^{137}}\mathtt{Cs}$ (used in radiotherapy) with half-life period $T = 30.1$ years and mass $m = \mathtt{team\, id \,number} \times 10^{-6}$ g each. Denote by $X_1,X_2,\dots,X_n$ the **i.i.d. r.v.**'s counting the number of decays in sample $i$ in one second.

1.  Specify the parameter of the Poisson distribution of $X_i$ (you'll need the atomic mass of *Cesium-137*)\
2.  Show that the distribution of the sample means of $X_1,\dots,X_n$ gets very close to a normal one as $n$ becomes large and identify that normal distribution. To this end,
    -   simulate the realization $x_1,x_2,\dots,x_n$ of the $X_i$ and calculate the sample mean $s=\overline{\mathbf{x}}$;
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of means and form the empirical cumulative distribution function $\hat F_{\mathbf{s}}$ of $\mathbf{s}$;
    -   identify $\mu$ and $\sigma^2$ such that the \textbf{c.d.f.} $F$ of $\mathscr{N}(\mu,\sigma^2)$ is close to the \textbf{e.c.d.f.} $\hat F_{\mathbf{s}}$ and plot both **c.d.f.**'s on one graph to visualize their proximity (use the proper scales!);
    -   calculate the maximal difference between the two \textbf{c.d.f.}'s;
    -   consider cases $n = 5$, $n = 10$, $n=50$ and comment on the results.\
3.  Calculate the largest possible value of $n$, for which the total number of decays in one second is less than $8 \times 10^8$ with probability at least $0.95$. To this end,
    -   obtain the theoretical bound on $n$ using Markov inequality, Chernoff bound and Central Limit Theorem, and compare the results;\
    -   simulate the realization $x_1,x_2,\dots,x_n$ of the $X_i$ and calculate the sum $s=x_1 + \cdots +x_n$;
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of sums;
    -   calculate the number of elements of the sample which are less than critical value ($8 \times 10^8$) and calculate the empirical probability; comment whether it is close to the desired level $0.95$

```{r}
# 1.1

m <- id/10^6            #
T <- 30.1*365*24*60*60  #
lambda <- log(2)/T      #  calculating everything we have to,
N <- 6*10**23*m/139     #  using the formulas above
mu <- N * lambda        #
K <- 1e3      
n <- 5



# X - r. v.
X <- rpois(n*K, lambda = mu)

# 2.1
sample_means <- colMeans(matrix(X, nrow=n))

sample_mean <- mean(sample_means)
```

```{r}

# 2.2
samp <- vector(mode = "numeric")

for (i in 1:K)
{
  samp <- colMeans(matrix(rpois(n*K, lambda = mu), nrow=n))
}
```

#### Next, calculate the parameters of the standard normal approximation

```{r}
# 2.3
mu <- mean(samp)
sigma <- sd(samp)
```

#### We can now plot ecdf and cdf

```{r}
xlims <- c(mu-3*sigma,mu+3*sigma)
Fs <- ecdf(samp)
plot(Fs, 
     xlim = xlims, 
     ylim = c(0,1),
     col = "blue",
     lwd = 2,
     main = "Comparison of ecdf and cdf")

curve(pnorm(x, mean = mu, sd = sigma), col = "red", lwd = 2, add = TRUE)
```

Calculation of the maximum difference between cdf an ecdf

```{r}
# 2.4

x = seq(mu-3*sigma, mu+3*sigma, by = 0.01)

max(abs(ecdf(samp)(x)-pnorm(x, mean = mean(samp), sd = sd(samp))))

```

```{r}
# 2.5
max_difference.func <- function(n) {
  means <- replicate(K, mean(colMeans(matrix(rpois(n*K, lambda = mu)))))
  norm_mu <- mean(means)
  sigma <- sd(means)
  
  x = seq(min(means), max(means), by = .01)
  return(max(abs(ecdf(means)(x)-pnorm(x, norm_mu, sigma))))
}

diff_5 <- max_difference.func(5)
diff_10 <- max_difference.func(10)
diff_50 <- max_difference.func(50)

diff_5
diff_10
diff_50


```

As we see, the number K and n are so large, that the average max difference are all the same. If we would take much bigger I think we will get better approximation. Anyway ecdf and cdf approximations are quite well.

```{r}
# 3.1
one_x <- colSums(matrix(rpois(n, lambda = mu), nrow=1))
spec_nums <- length(which(one_x>8*(10^8)))
cat("for n=5 the amound of sums that are bigger than 8e8\n")
spec_nums
cat("max sum\n")
max_sum = max(one_x)
max_sum
```

Markov inequality$$
P(S{n}>8e8) \leq \frac{1}{8e8}\mu n < .05
$$

(0.5 = 1-0.95)

(from the inequality express n) with $\mu\approx88256552$

$$
n< \frac{.05*8e8}{\mu}
$$

```{r}
cat(.05*8e8/N/lambda)
mu
sigma
```

as we see, this will occur for n \< .4532243

next we will obtain the upper bound of Chernoff inequality($\sigma\approx4295.206$):

$$
P(\frac{S_{n}-n\mu}{\sigma\sqrt{n}}> \frac{8e8-n\mu}{\sigma\sqrt{n}})\leq exp{{(-
\frac{8e8-n\mu}{4\sigma\sqrt(n)})^2}}
$$

I will equate it to .05 and find n with $\mu\approx88256438$ $\sigma\approx4365.243$

$$e^{-(\frac{8e8-n\mu}{4\sigma\sqrt(n)})^2} = .05 \Rightarrow ln(e^{-(\frac{8e8-n\mu}{4\sigma\sqrt(n)})^2}) = ln(.05)$$

$$n \approx 9.064$$

[wolfram](https://www.wolframalpha.com/input?i=-%28%28800000000-181846*4365.243%29%2F%284*4365.243*sqrt%28181846%29%29%29%5E2+%3D+ln%280.5%29)

### n = 9 (It is calculated using WolframAlpha)is an upper bound. The answer is pretty precise and already satisfies the condition

Central limit theorem

$$Z_n:=\frac{S_n-\mu\cdot n}{\sigma\sqrt{n}}$$ $$Z_n\sim N(0, 1)$$ $$P(\frac{S_{n}-n\mu}{\sigma\sqrt{n}}< \frac{8\cdot 10^8-n\mu}{\sigma\sqrt{n}}) \rightarrow \Phi(\frac{8\cdot 10^8-n\mu}{\sigma\sqrt{n}})\geq .95$$

$$\frac{8\cdot 10^8-n\mu}{\sigma\sqrt{n}} = 1,65$$

[new_wolfram](https://www.wolframalpha.com/input?i=%28800000000-n*88256438%29%2F%284365.243*sqrt%28n%29%29%3D1.65)

The result is the same $n \approx 9.064$

### n = 9 (It is calculated using WolframAlpha)is an upper bound. The answer is pretty precise and already satisfies the condition

compare the result

This satisfies the results

```{r}
n <- 9
colum_sum <- colSums(matrix(rpois(n * K, lambda = mu), nrow = n))
condit <- length(which(colum_sum > 8e8))
print(condit)

```

This does not satisfies the result

```{r}
n <- 18
colum_sum <- colSums(matrix(rpois(n * K, lambda = mu), nrow = n))
condit <- length(which(colum_sum > 8e8))
print(condit)

```

Solving these examples, we were able to apply previously learned knowledge about different types of distributions, as well as verify the correctness of how these distributions work with real cases. We were also able to verify the effectiveness of Markov's inequality, Chernoff bound, Central Limit Theorem and much more


### Task 3 (Taras Yaroshko)

#### In this task, we use the Central Limit Theorem approximation for continuous random variables.x

#### One of the devices to measure radioactivity level at a given location is the Geiger counter. When the radioactive level is almost constant, the time between two consecutive clicks of the Geiger counter is an exponentially distributed random variable with parameter $\nu_1 = 38$. Denote by $X_k$ the random time between the $(k-1)^{\mathrm{st}}$ and $k^{\mathrm{th}}$ click of the counter.

1.  Show that the distribution of the sample means of $X_1, X_2,\dots,X_n$ gets very close to a normal one (which one?) as $n$ becomes large. To this end,
    -   simulate the realizations $x_1,x_2,\dots,x_n$ of the \textbf{r.v.} $X_i$ and calculate the sample mean $s=\overline{\mathbf{x}}$;\
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of means and then the \emph{empirical cumulative distribution} function $F_{\mathbf{s}}$ of $\mathbf{s}$;\
    -   identify $\mu$ and $\sigma^2$ such that the \textbf{c.d.f.} of $\mathscr{N}(\mu,\sigma^2)$ is close to the \textbf{e.c.d.f.} $F_{\mathbf{s}}$ of and plot both \textbf{c.d.f.}'s on one graph to visualize their proximity;\
    -   calculate the maximal difference between the two \textbf{c.d.f.}'s;\
    -   consider cases $n = 5$, $n = 10$, $n=50$ and comment on the results.
2.  The place can be considered safe when the number of clicks in one minute does not exceed $100$. It is known that the parameter $\nu$ of the resulting exponential distribution is proportional to the number $N$ of the radioactive samples, i.e., $\nu = \nu_1*N$, where $\nu_1$ is the parameter for one sample. Determine the maximal number of radioactive samples that can be stored in that place so that, with probability $0.95$, the place is identified as safe. To do this,
    -   express the event of interest in terms of the \textbf{r.v.} $S:= X_1 + \cdots + X_{100}$;\
    -   obtain the theoretical bounds on $N$ using the Markov inequality, Chebyshev bound and Central Limit Theorem and compare the results;\
    -   with the predicted $N$ and thus $\nu$, simulate the realization $x_1,x_2,\dots,x_{100}$ of the $X_i$ and of the sum $S = X_1 + \cdots + X_{100}$;\
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of total times until the $100^{\mathrm{th}}$ click;\
    -   estimate the probability that the location is identified as safe and compare to the desired level $0.95$

### 3.1

#### Generating samples an sample means:

```{r}
nu1 <- 38
count <- 1e3
n <- 100
sample_means <- colMeans(matrix(rexp(n*count, rate = nu1), nrow=n))
```

#### Calculating the parameters of the standard normal approximation:

```{r}
# mu <- mean(sample_means)
# sigma <- sd(sample_means)
mu <- 1 / nu1
sigma <- mu / sqrt(n)
```

#### We can now plot ecdf and cdf:

```{r}
xlims <- c(mu-3*sigma,mu+3*sigma)
ylims <- c(-0.5, 1.5)
Fs <- ecdf(sample_means)
plot(Fs, xlim = xlims, ylim=ylims, col = "blue", lwd = 1, main = "Comparison of ecdf and cdf")
curve(pnorm(x, mean = mu, sd = sigma), col = "red", lwd = 2, add = TRUE)
```

#### Calculating the maximal difference between the two cdf's:

```{r}
max(abs(Fs(xlims) - pnorm(xlims, mean = mu, sd = sigma)))
```

### 3.2

#### The event of interest:

```{r}
num = 100
S = sum(sample_means[1:100])
S
```

#### Theoretical bounds:

**Markov bound:**

```{r}
seconds = 1
safe_perc = 0.95
#Upper bound
N_Markov = num / (nu1 * seconds * safe_perc)
N_Markov
```

**Chebyshev's bound:**

```{r}
N_Chebyshev = (num * safe_perc + sqrt(num * safe_perc - safe_perc * safe_perc)) / (nu1 * seconds * safe_perc)
N_Chebyshev
```

**Central Limit Theorem:**

```{r}
N_CLT = 1
while(TRUE){
  prob = pnorm((seconds - (num / (nu1 * N_CLT))) / (sqrt(num) / (nu1 * N_CLT)))
  if (prob >= 0.05){
    break
  }
  N_CLT = N_CLT + 1
}
N_CLT = N_CLT - 1
N_CLT
```

**Simulate the realization**

```{r}
s_after = sum(rexp(100, rate = N_CLT * nu1))
s_after
```

**Repeat it K times**

```{r}
sums = colSums(matrix(rexp(100*count, rate = N_CLT * nu1), nrow = 100))
```

**Estimate the probability that the location is identified as safe**

```{r}
mean(sums >= 1)
```

### Conclusions

In this task we've learned how the exponential r.v can be used to approximate the outcome of a real-world experiment, namely the amount of particles that decay per unit of time and how we can use that to find out whether the place is safe to be in. Also, we've noticed how R can come in handy in this task by providing powerful and expressive syntax to perform difficult tasks in 1 line of code.
